Measuring the double-drift illusion and its resets with hand trajectories

### 2022-01-20 Manuscript.Rmd
added stats for the binned fits, and references for the ex-Gaussian distribution

---
title: "Measuring the double-drift illusion and its resets with hand trajectories"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document:
    reference_docx: template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{amsmath}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

While the marginal distribution plot in Fig 5A qualitatively appear an equally good fit to the data, the better fit for the distribution on the time of resets could be explained be a better generalization throughout the data set. We test this by dividing the data into segments, each with an equal number of resets, based on variables orthogonal to the fitted variable (Y-coordinates for the spatial offset of resets and angle for reset time). The median for each segment of data is shown in Fig 5B,C. We fit a gamma distribution to each segment, then use the data points outside the segment to get a log likelihood as a measure of generalization and sum them (spatial: XXXXX, temporal: XXXXX). We use this to calculate AICs and a relative likelihood (spatial: XXXXX, temporal: XXXXX, _p_<XXXXX) which shows that a gamma distribution fit to reset times in one segment of the data generalizes better to the rest of the data, as compared to a gamma distribution fit to the spatial offset of resets.

@@

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in their experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe [-@Nakayama2020] claim that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the distribution of resets throughout time are largely in line with a mechanism based on random distractions of attention, which may involve eye-movement planning and execution. The rate of resets that we found here, ~0.75 / _s_, is too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements. On the other hand, gamma distributions [e.g. @McGill1965; @Wolfe2002] do model a random occurrence of events (rate parameter) where only some fraction of these trigger an action (shape parameter). In other words, the good fit of a gamma distribution may reveal some properties of the process underlying resets of the double-drift illusion. While the process giving rise to resets is likely memoryless, a process giving rise to a gamma distribution (with shape > 1) is not, it does capture two ideas: 1) resets are triggered by random events (such as distractions of attention or microsaccades), and 2) only some of these actually lead to a reset. An alternative to the gamma distribution, the ex-Gaussian (ref), could be a more realistic alternative but does not fit the data as well. However, the process that actually leads to resets remains to be seen in future work.

>>>>

---
title: "Measuring the double-drift illusion and its resets with hand trajectories"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2
output:
  word_document:
    reference_docx: template.docx
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{amsmath}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

While the marginal distribution plot in Fig 5A qualitatively appear an equally good fit to the data, the better fit for the distribution on the time of resets could be explained be a better generalization throughout the data set. We test this by dividing the data into segments, each with an equal number of resets, based on variables orthogonal to the fitted variable (Y-coordinates for the spatial offset of resets and angle for reset time). The median for each segment of data is shown in Fig 5B,C. We fit a gamma distribution to each segment, then use the data points outside the segment to get a log likelihood as a measure of generalization and sum them (spatial: -3087.9, temporal: -1806.3). We use this to calculate AICs and a relative likelihood (spatial: 6191.7, temporal: 3628.6, _p_<.001) which shows that a gamma distribution fit to reset times in one segment of the data generalizes better to the rest of the data, as compared to a gamma distribution fit to the spatial offset of resets.

@@

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in their experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe [-@Nakayama2020] claim that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the distribution of resets throughout time are largely in line with a mechanism based on random distractions of attention, which may involve eye-movement planning and execution. The rate of resets that we found here, ~0.75 / _s_, is too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements. On the other hand, gamma distributions [e.g. @McGill1965; @Wolfe2002] do model a random occurrence of events (rate parameter) where only some fraction of these trigger an action (shape parameter). In other words, the good fit of a gamma distribution may reveal some properties of the process underlying resets of the double-drift illusion. While the process giving rise to resets is likely memoryless, a process giving rise to a gamma distribution (with shape > 1) is not, it does capture two ideas: 1) resets are triggered by random events (such as distractions of attention or microsaccades), and 2) only some of these actually lead to a reset. An alternative to the gamma distribution, the ex-Gaussian [e.g. @Matzke2009; @Guy2020], could be a more realistic alternative but does not fit the data as well. However, the process that actually leads to resets remains to be seen in future work.
